---
title: "Introduction-to-rPandas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction-to-rPandas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(rPandas)
```


## Introduction
rPandas is an R package designed to serve as a translation layer, allowing R users to leverage the power and speed of the Python pandas library without ever leaving their R session.

For R users familiar with dplyr or data.table, the rPandas syntax will feel natural. The package provides a set of R functions (e.g., rp_filter(), rp_select()) that:

Capture R code (like carat > 1).

Translate it into an equivalent Python/pandas command string.

Execute the Python code in the background via the reticulate package.

Return the resulting data.frame back to your R session.

This allows you to write R code while the data processing is handled by Python's pandas library.

## Setup and Installation
### Python Dependencies
rPandas depends on a working Python installation and the pandas library. The reticulate package handles the R‑to‑Python connection.

### Quick installation (default environment)
You can install pandas into a dedicated R environment by running:

```{r eval=FALSE}
reticulate::py_install("pandas")
```
#### Troubleshooting the Python environment

If you encounter issues (e.g., “pandas not found”), rPandas provides a built‑in health check:

```{r eval=FALSE}
# Run this if you have connection issues
rp_check_env()
```

The health check prints the Python path reticulate is using and whether `pandas` is installed.

##### Finding and selecting the correct Python environment

You can list all available conda environments (if you use Conda) with:

```{r eval=FALSE}
reticulate::conda_list()
```

Once you identify the desired environment (e.g., the first one), you can tell reticulate to use it **before loading rPandas**:

```{r eval=FALSE}
# Replace with the path from conda_list()
reticulate::use_python(python = reticulate::conda_list()$python[1], required = TRUE)

# Or, if you prefer to use a conda environment by name:
reticulate::use_condaenv("your_environment_name", required = TRUE)
```

For system Python installations, simply provide the path to the Python binary:

```{r eval=FALSE}
reticulate::use_python("/usr/local/bin/python3", required = TRUE)
```

After setting the environment, rerun `rp_check_env()` to confirm everything is ready. For more detailed guidance, see the [reticulate documentation](https://rstudio.github.io/reticulate/).

> **Note:** The code chunks above are not executed when building the vignette (they are meant to be run interactively by the user).

## The Core Verbs
All examples will use the diamonds dataset from the ggplot2 package.

```{r}
# Make sure ggplot2 is installed to access the data
data(diamonds, package = "ggplot2")
head(diamonds)
```
**1. rp_filter():**
This verb filters rows based on a logical expression, similar to dplyr::filter().
It supports common logical operators: & (AND), | (OR), ! (NOT), %in%, and %notin%

 ``` {r} 
# Simple condition
v1 <- rp_filter(diamonds, carat > 1)
print(head(v1))

# AND: multiple conditions
v2 <- rp_filter(diamonds, carat > 1 & cut == "Ideal")
print(head(v2))

# OR: use | (pipe)
v3 <- rp_filter(diamonds, color == "D" | color == "E")
print(head(v3))

# NOT: use !
v4 <- rp_filter(diamonds, !(price > 10000))
print(head(v4))

# %in% operator
v5 <- rp_filter(diamonds, color %in% c("D", "E", "F"))
print(head(v5))

# %notin% (if implemented)
v6 <- rp_filter(diamonds, color %notin% c("D", "E", "F"))
print(head(v6))
 ```

**2. rp_select():**
This verb selects specific columns by name, similar to dplyr::select().

```{r}
# Select three columns
v4 <- rp_select(diamonds, carat, cut, price)
print(head(v4))
```

**3. rp_sort():**
This verb sorts the data frame by one or more columns, similar to dplyr::arrange().
```{r}
# Sort by price (ascending by default)
v8 <- rp_sort(diamonds, price)
print(head(v8))

# Use desc() to sort in descending order
v9 <- rp_sort(diamonds, desc(price))
print(head(v9))

# Sort by multiple columns
v10 <- rp_sort(diamonds, cut, desc(price))
print(head(v10))
```


**4. rp_mutate():**
This verb creates new columns or modifies existing ones, similar to dplyr::mutate().
You can also remove columns using the to_remove argument.
```{r}

# Create a new column
v11 <- rp_mutate(diamonds, price_per_carat = price / carat)
print(head(v11))

# Create multiple columns
v12 <- rp_mutate(
  diamonds, 
  price_per_carat = price / carat,
  depth_pct = depth / 100
)
print(head(v12))

# Remove one or more columns
v13 <- rp_mutate(diamonds, to_remove = c("table", "depth"))
print(head(v13))
```


**5. rp_summarize():**
This verb collapses a data frame into a summary, often after grouping. It is similar to dplyr::summarise().
```{r}

# Summarize the entire data frame
v14 <- rp_summarize(diamonds, avg_price = mean(price), max_carat = max(carat))
print(v14)

# Group by one column (unquoted)
v15 <- rp_summarize(diamonds, avg_price = mean(price), .by = cut)
print(v15)

# Group by multiple columns (unquoted)
v16 <- rp_summarize(
  diamonds, 
  avg_price = mean(price), 
  count = n(),
  .by = c(cut, color)
)
print(head(v16))

# Grouping also accepts quoted column names
v17 <- rp_summarize(
  diamonds, 
  avg_price = mean(price), 
  .by = c("cut", "color")
)
print(head(v17))
```

**6. rp_calculate()– Apply Multiple Functions to Multiple Columns**
This verb is a powerful alternative to rp_summarize. It applies the same set of functions (e.g., "mean", "sd") to every selected column.
```{r}

# Apply two functions to two columns, grouped by 'cut'
v13 <- rp_calculate(
  diamonds,
  price, carat,
  the.functions = c("mean", "sd"),
  .by = cut
)
print(head(v13))
```

**7. rp_first_k_rows() and rp_last_k_rows() – First/Last K Rows per Group**
These verbs extract the first or last k rows from the whole data frame, or from each group when .by is provided.
```{r}
# First 3 rows overall
v19 <- rp_first_k_rows(diamonds, k = 3)
print(v19)

# Last 2 rows per group (cut and clarity)
v20 <- rp_last_k_rows(diamonds, k = 2, .by = c(cut, clarity))
print(head(v20))

# Both quoted and unquoted group specifications work
v21 <- rp_first_k_rows(diamonds, k = 1, .by = c("cut", "color"))
print(v21)
```

**8. rp_count() – Count Rows (Overall or by Group)**
This verb returns the number of rows in the data frame, optionally by groups.
```{r}
# Total row count
v22 <- rp_count(diamonds)
print(v22)

# Count per group
v23 <- rp_count(diamonds, .by = cut)
print(v23)

# Count per combination of multiple groups
v24 <- rp_count(diamonds, .by = c(cut, color))
print(head(v24))
```

## Chaining Verbs with the Pipe
All rPandas functions are "pipe-friendly" and use .data as their first argument, allowing you to chain operations together using magrittr's %>% pipe (or the native R |>).

```{r}
# Load the pipe
v25 <- diamonds |>
  rp_filter(carat > 1 & color == "D") |>
  rp_mutate(price_per_carat = price / carat) |>
  rp_summarize(avg_ppc = mean(price_per_carat), .by = cut) |>
  rp_sort(desc(avg_ppc))

print(head(v25))
```

## Viewing the Python Code
A key feature of rPandas is its role as a learning and translation tool. Every user-facing verb has a return.as argument that lets you inspect the Python code it generates.

You can set return.as to:

**"result" (default):** Returns the final R data frame.

**"code":** Returns the generated Python command as a string.

**"all":** Returns a list containing both the result and the code.

```{r}
# See the code for a simple filter
rp_filter(diamonds, carat > 1 & price < 400, return.as = "code")

# See the code for a mutate
rp_mutate(diamonds, ppc = price / carat, return.as = "code")

# See the code for a complex summary
rp_summarize(
  diamonds,
  avg_price = mean(price), 
  count = n(),
  .by = c(cut, color),
  return.as = "code"
)
```

## Customizing the Generated Code with table_name
By default, when you request return.as = "code", the generated Python code uses the placeholder rpandas_df_in as the DataFrame name. This is because the function only receives the 
data object, not its name. However, you can provide a custom name using the table_name argument (available in all verbs). This is especially useful when you want to copy‑paste the 
code into a Python script or notebook.
```{r}
# Default placeholder
rp_filter(diamonds, carat > 1, return.as = "code")

# With custom table name
rp_filter(diamonds, carat > 1, table_name = "diamonds", return.as = "code")
```
The output changes from:
**"rpandas_df_in.query('(carat > 1)')"** to: **"diamonds.query('(carat > 1)')"**.
This works for all verbs – rp_select, rp_mutate, rp_summarize, etc. Simply pass table_name = "your_data_frame_name" as an argument.